---
title: "Importing WaPo Dataset on Nonprofits with Disclosed Material Mistatements or Diversions"
output: html_notebook
params:
  filingyear:
    label: "990 Filing Year"
    value: 2011
    input: slider
    min: 2011
    max: 2015
    step: 1
---

```{r include=FALSE}
# Begin by loading the required packages.
library(plyr)
library(dplyr)
library(rvest)
library(RCurl)
library(XML)
```

In the interest of time, let's only evaluate pull 990 filings for those organizations listed in the Washington Post dataset of having reported material mistatements or diversions. http://www.washingtonpost.com/wp-srv/special/local/nonprofit-diversions-database/ 

For code development and exploratory purposes only, we will subset the WaPo df to include only those organizations that reported material diversions in the 2010 tax year (and would therefore be included in the 2011 filings). We will then compare this list to the IRS index file for 2011, creating a new df with the 990 URLs for only these organizations.

```{r}

filingyear <- "2011"

library(readr)
WaPo_Dataset_on_Material_Diversions_Disclosed_in_990 <- read_csv("~/Box Sync/Harvard University/Forensic Accounting and Fraud Examination/Research Paper/WaPo Dataset on Material Diversions Disclosed in 990.csv",
                                                                 col_types = cols(`Revenue ` = col_number()))

# Force WaPo df to uppercase text only (in order to merge later)
WaPo_Dataset_on_Material_Diversions_Disclosed_in_990 <-
  mutate_all(WaPo_Dataset_on_Material_Diversions_Disclosed_in_990, funs(toupper))

# Rename Organization column in WaPo dataset to match the IRS index files (also required to merge).
names(WaPo_Dataset_on_Material_Diversions_Disclosed_in_990)
names(irs2011)
names(WaPo_Dataset_on_Material_Diversions_Disclosed_in_990)[1] <- names(irs2011)[6]

# Subset for filing year
WaPo_subset <- subset(WaPo_Dataset_on_Material_Diversions_Disclosed_in_990,
                      WaPo_Dataset_on_Material_Diversions_Disclosed_in_990$`Tax Year ` == filingyear)

diversionsonly_2011 <- merge(WaPo_subset, 
                             irs2011, 
                             by = "TAXPAYER_NAME")
```

The code above starts us on the right direction, but let's update the "hard coded" filing year to something  parameterized. Referencing the video tutorial at https://www.rstudio.com/resources/webinars/beyond-static-reports-with-r-markdown/. 

Note: This code will only work if each of the IRS XML files have an identical (or nearly identical) return type. We will hardcode the return type to "990", but other possible options are 990EO, 990EZ, and 990O.
  Note: If you change the return type, you will also need to change the XML path in the for-loop.

```{r}

library(readr)
WaPo_Dataset_on_Material_Diversions_Disclosed_in_990 <- read_csv("~/Box Sync/Harvard University/Forensic Accounting and Fraud Examination/Research Paper/WaPo Dataset on Material Diversions Disclosed in 990.csv",
                                                                 col_types = cols(`Revenue ` = col_number()))

# Force WaPo df to uppercase text only (in order to merge later)
WaPo_Dataset_on_Material_Diversions_Disclosed_in_990 <-
  mutate_all(WaPo_Dataset_on_Material_Diversions_Disclosed_in_990, funs(toupper))

# Rename Organization column in WaPo dataset to match the IRS index files (also required to merge).
names(WaPo_Dataset_on_Material_Diversions_Disclosed_in_990)
names(irs2011)
names(WaPo_Dataset_on_Material_Diversions_Disclosed_in_990)[1] <- names(irs2011)[6]

# Subset WaPo for filing year
WaPo_subset <- subset(WaPo_Dataset_on_Material_Diversions_Disclosed_in_990,
                      WaPo_Dataset_on_Material_Diversions_Disclosed_in_990$`Tax Year ` == params$filingyear)

# Subset IRS for return type
irs2011_subset <- subset(irs2011, 
                         irs2011$RETURN_TYPE == "990")

diversionsonly_2011 <- merge(WaPo_subset, 
                             irs2011_subset, 
                             by = "TAXPAYER_NAME")

# Remove duplicate URLs
diversionsonly_2011 <- 
  diversionsonly_2011 %>%
  distinct(URL)
```

